---
title: "gluster Demonstration"
author: Ruby Xiong and Simon Vandekar
format: html
editor: visual
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Loading Cody's Data

```{r}
library(ggplot2)
cell = readRDS('/media/disk2/atlas_mxif/data/threshold_quantification_global_20230315.rds')
#allMarkers = grep('^Median_Cell_', names(cell), value=TRUE)
allMarkers <- c("CD3D", "CD4", "CD8", "CD68", "CD20", "CD11B", "FOXP3", "LYSOZYME", "DAPI")
allMarkers = paste0("Median_Cell_", allMarkers)
cell$slide_id = cell$SlideID

# compute silver standard
at = grep('Area_Thresh_*', names(cell), value=TRUE)
phenotypes = gsub('Area_Thresh_', '', at) 
cell[,phenotypes] = sweep(cell[,at], 1, cell$Cell_Area, FUN = '/')>0.3
```

## Loading Julia's mIF data

```{r, message=FALSE, eval=FALSE}
library(VectraPolarisData)
spe_lung <- HumanLungCancerV3()

# get the cell level imaging data with spatial coordinates
cell = cbind(as.data.frame(colData(spe_lung)), as.data.frame(spatialCoords(spe_lung)) )

#devtools::install_github('statimagcoll/gluster')
allMarkers = grep('^entire_cell_.*total$', names(cell), value=TRUE)

# get the clinical data for each sample
#cd = as.data.frame(metadata(spe_lung)$clinical_data)
```

## Running gluster on a single slide

Before fitting the model, we apply a basic normalization and set some boundary constraints. If constraints aren't known, we can set all the constraints to `NULL`.

```{r, setup}
library(gluster)
# normalize the data 
nzNormedMarkers = paste0('nzNorm_',allMarkers)
cell = do.call(rbind, lapply(split(cell, cell$slide_id), function(df){df[,nzNormedMarkers] = log10(1+sweep(df[,allMarkers], 2, colMeans(df[,allMarkers], na.rm = TRUE ), FUN = '/' ) ); df }) ) # replace(df[,allMarkers], df[,allMarkers]==0, NA)


quantileBoundaries = list(matrix(c(0,.8, .9, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          NULL,
                          NULL,
                          NULL,
                          NULL)
boundaries = list(matrix(c(0,.3, .5, 1), byrow = TRUE, nrow=2),
                  NULL,
                  NULL,
                  NULL,
                  matrix(c(0,.3, .45, Inf), byrow = TRUE, nrow=2),
                  matrix(c(0,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(0,.4, .45, Inf), byrow = TRUE, nrow=2),
                  NULL,
                  NULL)
names(quantileBoundaries) = names(boundaries) = nzNormedMarkers
```

The code below fits the model in the first slide.

```{r singleFit, eval=FALSE}

# split the data frame
cells = cell[cell$slide_id==unique(cell$slide_id)[2],]
#unconstrGMMfit = gluster(cells[,nzNormedMarkers])
constrGMMfit = gluster(cells[,nzNormedMarkers], boundaryMarkers = boundaries,
                 qboundaryMarkers=quantileBoundaries)
```

### Quality checking model fit

```{r, visualizeOneSlide, eval=FALSE}
# check convergence
sapply(constrGMMfit$fit, function (x) x$convergence)
# visualize constrained fits with quantile drawn on histogram
invisible(sapply(nzNormedMarkers, function(x) plot(constrGMMfit, marker = x, boundary = quantile(constrGMMfit$expressionX[,x], probs=quantileBoundaries[[x]][2,1], na.rm=TRUE), title=x ) ))
```

Convergence failed for the CD8 marker. Adjusting the constraints might improve the fit. We add in a lower bound of 1/2 for the expressed cell population for CD8. Then we refit the data with new constraints for all the markers based on visualization of previous histograms. If it continues to fail, we can deal with it later by obtaining an estimated fit from the other slide models.

```{r}
# adjusted boundaries
quantileBoundaries2 = list(matrix(c(0,.8, .9, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          NULL,
                          NULL,
                          NULL,
                          NULL)
boundaries2 = list(NULL,
                  matrix(c(-Inf,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .75, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .75, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.4, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.4, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.4, .4, Inf), byrow = TRUE, nrow=2))
names(quantileBoundaries2) = names(boundaries2) = nzNormedMarkers
#constrGMMfit2 = gluster(cells[,nzNormedMarkers], boundaryMarkers = boundaries2,
#                 qboundaryMarkers=quantileBoundaries2)
```

We can contrast the two model fits with the `hist_sr_contrast` function. This shows the original fit in green, with the updated fit in red to see how the histogram has changed between the two.

```{r, eval=FALSE}
# comparing constrained and unconstrained fits
invisible(sapply(nzNormedMarkers, function(x) hist_sr_constrast(constrGMMfit, constrGMMfit2, marker=x, title=x) ))
```

## Running with sub-batches

Sub-batches can be used if there are multiple regions per slide to assess the fit within each region. The model is fit at the slide level and then model fit is assessed by batch.

```{r subBatch, eval=FALSE}
# fitting on the same slide and getting model fit results by batch
batchConstrGMMfit = gluster(cells[[1]][,nzNormedMarkers], boundaryMarkers = boundaries,
                 qboundaryMarkers=quantileBoundaries, subBatch = cells[[1]]$sample_id)
# visualize by subBatch for the first markers
trash = sapply(unique(cells[[1]]$sample_id), function(x) plot.gluster(batchConstrGMMfit, subBatch = x, title=x) )
```

## Running gluster on multiple slides

We can use `groupGluster` to run gluster on multiple slides. We start with the second round of boundaries that we chose from the first batch and run it on the same number of cores as slides. Each slide will be assigned to a separate core. The same boundaries are used for all the slides.

```{r, batchFit, eval=FALSE}
constrCfGMMbunch = groupGluster(cell[,nzNormedMarkers], slide=cell$slide_id, boundaryMarkers=boundaries2, qboundaryMarkers = quantileBoundaries2, n.cores = 12)
saveRDS(constrCfGMMbunch, file='/media/disk2/atlas_mxif/data/cfGMM_cody_fit.rds')


# debugging gluster
# for(i in unique(cell$slide_id)){
#   cat(i,'\n')
#   test = gluster(cell[cell$slide_id==i,nzNormedMarkers], boundaryMarkers=boundaries2, qboundaryMarkers = quantileBoundaries2)
# }

# library(reactable)
# convergenceTab = sapply(constrCfGMMbunch, function(x) sapply(x$fit, function(y){ y$convergence}))
# # the model converged for most slides
# rowMeans(convergenceTab)
# reactable(convergenceTab)


```


# Visualizing the fitted distributions {.tabset}


```{r, results='asis', fig.width=8, fig.height=3, warning=FALSE}
constrCfGMMbunch = readRDS('/media/disk2/atlas_mxif/data/cfGMM_cody_fit.rds')
for(marker in names(boundaries)){
  cat('\n## ', marker, '\n')
  plot(constrCfGMMbunch, marker=marker)
  invisible(capture.output(sapply(names(constrCfGMMbunch), function(x) plot(constrCfGMMbunch[[x]], marker = marker, boundary = quantile(constrCfGMMbunch[[marker]]$expressionX[,1], probs=quantileBoundaries[[marker]][2,1]), title = x, tabl=TRUE ) )))
  cat('\n\n')
}
```


# Adjusting the boundaries


We are unlikely to get a biologically consistent model fit without applying constraints or evaluating the boundaries. To improve fit across all slides, we visually assess the fit and adjust our initial starting boundaries. This can be done by visualizing each histogram (time intensive) or visualizing the estimated parameters across all slides and assessing the fit.
After visualizing the fits we can adjust the boundaries. FOXP3, is it a sparsely expressed or continuous marker?

```{r adjustedBoundaries}
quantileBoundaries3 = list(matrix(c(0,.8, .9, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          NULL,
                          NULL,
                          NULL)
# might look more like the manual labels
boundaries3 = list(matrix(c(-Inf,.3, .8, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .6, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf, 0, .5, 1.5), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .45, 1), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .4, .75), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.4, .45, Inf), byrow = TRUE, nrow=2),
                  NULL)

# better fit to distribution
boundaries4 = list(matrix(c(-Inf,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .6, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf, 0, .5, 1.5), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .45, 1), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.3, .4, .75), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.4, .45, Inf), byrow = TRUE, nrow=2),
                  matrix(c(-Inf,.4, .4, Inf), byrow = TRUE, nrow=2))
```


```{r, eval=FALSE}
constrCfGMMbunch = groupGluster(cell[,nzNormedMarkers], slide=cell$slide_id, boundaryMarkers=boundaries3, qboundaryMarkers = quantileBoundaries3, n.cores = 12)
saveRDS(constrCfGMMbunch, file='/media/disk2/atlas_mxif/data/cfGMM_cody_fit2.rds')

constrCfGMMbunch = groupGluster(cell[,nzNormedMarkers], slide=cell$slide_id, boundaryMarkers=boundaries4, qboundaryMarkers = quantileBoundaries3, n.cores = 12)
saveRDS(constrCfGMMbunch, file='/media/disk2/atlas_mxif/data/cfGMM_cody_fit3.rds')
```


# The new fitted distributions {.tabset}

```{r, results='asis', fig.width=8, fig.height=3, warning=FALSE}
constrCfGMMbunch = readRDS('/media/disk2/atlas_mxif/data/cfGMM_cody_fit3.rds')
marker = 2
for(marker in names(boundaries)){
  cat('\n## ', marker, '\n')
  plot(constrCfGMMbunch, marker=marker)
  invisible(capture.output(sapply(names(constrCfGMMbunch), function(x) plot(constrCfGMMbunch[[x]], marker = marker, title = x, tabl=TRUE ) ))) # boundary = quantile(constrCfGMMbunch[[marker]]$expressionX[,1], probs=quantileBoundaries[[marker]][2,1]), 
  cat('\n\n')
}
```


# Evaluate fit using Cohen's Kappa

```{r, fig.width=5, fig.height=5}
#constrCfGMMbunch = readRDS('/media/disk2/atlas_mxif/data/cfGMM_cody_fit3.rds')
# constrCfGMMbunch = lapply(constrCfGMMbunch, function(x, subBatch=1){
#   result = as.data.frame(do.call(cbind, lapply(names(x$expressionX), function(nm, subBatch){
#   pars = x$params[[nm]][[subBatch]]
#   pgamma(x$expressionX[[nm]], shape=pars[2,3],scale=pars[3,3])
# }, subBatch=subBatch )))
#   names(result) = names(x$expressionX)
#   x$expressionW = result
#   return(x)
#   } )
# class(constrCfGMMbunch) = c('groupGluster', class(constrCfGMMbunch))

# Assumes cell was grouped by slideID
merged = unsplit(lapply(names(constrCfGMMbunch), function(x){
  constrCfGMMbunch[[x]]$expressionZ<-as.data.frame(constrCfGMMbunch[[x]]$expressionZ>0.5)
  names(constrCfGMMbunch[[x]]$expressionZ) = paste0('cfGMMPheno_', names(constrCfGMMbunch[[x]]$expressionZ))
  rownames(constrCfGMMbunch[[x]]$expressionZ) = paste0(x, rownames(constrCfGMMbunch[[x]]$expressionZ))
  constrCfGMMbunch[[x]]$expressionZ}), cell$slide_id)
cell[,names(merged)] = merged

merged = unsplit(lapply(names(constrCfGMMbunch), function(x){
  constrCfGMMbunch[[x]]$expressionZ<-as.data.frame(constrCfGMMbunch[[x]]$expressionW>0.5)
  names(constrCfGMMbunch[[x]]$expressionW) = paste0('cfGMMPhenoMarg_', names(constrCfGMMbunch[[x]]$expressionW))
  rownames(constrCfGMMbunch[[x]]$expressionW) = paste0(x, rownames(constrCfGMMbunch[[x]]$expressionW))
  constrCfGMMbunch[[x]]$expressionW}), cell$slide_id)
cell[,names(merged)] = merged

saveRDS(cell, file='/media/disk2/atlas_mxif/data/cody_cfGMM.rds')

invisible(kappaGroupGluster(posterior = cell[, paste0('cfGMMPheno_nzNorm_Median_Cell_', phenotypes)],
                            marginal=cell[, paste0('cfGMMPhenoMarg_nzNorm_Median_Cell_', phenotypes)],
                            standard = cell[,phenotypes], batch = cell$slide_id))
# 

# computes cell proportions for each phenotyping method
res = do.call(rbind, by(cell, cell$slide_id, function(df) colMeans(df[,c(phenotypes, paste0('cfGMMPheno_nzNorm_Median_Cell_', phenotypes) , paste0('cfGMMPhenoMarg_nzNorm_Median_Cell_', phenotypes))], na.rm=TRUE) ))
# plots them
invisible(lapply(phenotypes, function(ph){
  plot(res[,ph], res[,paste0('cfGMMPheno_nzNorm_Median_Cell_', ph)], type='n', bty='l', xlab='Manual', ylab='cfGMM', main=ph)
  text(res[,ph], res[,paste0('cfGMMPheno_nzNorm_Median_Cell_', ph)], rownames(res), cex=0.5)
  text(res[,ph], res[,paste0('cfGMMPhenoMarg_nzNorm_Median_Cell_', ph)], rownames(res), cex=0.5, col='red')
  abline(a=0, b=1)
} ) )
```

