---
title: "gluster Demonstration"
author: Ruby Xiong and Simon Vandekar
format: html
editor: visual
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Loading Julia's mIF data

```{r, message=FALSE}
library(VectraPolarisData)
spe_lung <- HumanLungCancerV3()

# get the cell level imaging data with spatial coordinates
cell = cbind(as.data.frame(colData(spe_lung)), as.data.frame(spatialCoords(spe_lung)) )

# get the clinical data for each sample
#cd = as.data.frame(metadata(spe_lung)$clinical_data)
```

## Running gluster on a single slide

Before fitting the model, we apply a basic normalization and set some boundary constraints. If constraints aren't known, we can set all the constraints to `NULL`.

```{r, setup}
library(gluster)
#devtools::install_github('statimagcoll/gluster')
allMarkers = grep('^entire_cell_.*total$', names(cell), value=TRUE)
# normalize the data 
nzNormedMarkers = paste0('nzNorm_',allMarkers)
cell = do.call(rbind, lapply(split(cell, cell$slide_id), function(df){df[,nzNormedMarkers] = log10(1+sweep(df[,allMarkers], 2, colMeans(replace(df[,allMarkers], df[,allMarkers]==0, NA), na.rm = TRUE ), FUN = '/' ) ); df }) )


quantileBoundaries = list(matrix(c(0,.8, .9, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          NULL,
                          NULL,
                          NULL)
boundaries = list(matrix(c(0,.3, .5, 1), byrow = TRUE, nrow=2),
                  NULL,
                  NULL,
                  NULL,
                  matrix(c(0,.3, .45, Inf), byrow = TRUE, nrow=2),
                  matrix(c(0,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(0,.4, .45, Inf), byrow = TRUE, nrow=2),
                  NULL)
names(quantileBoundaries) = names(boundaries) = nzNormedMarkers
```

The code below fits the model in the first slide.

```{r singleFit}

# split the data frame
cells = split(cell, cell$slide_id)
unconstrGMMfit = gluster(cells[[1]][,nzNormedMarkers])
constrGMMfit = gluster(cells[[1]][,nzNormedMarkers], boundaryMarkers = boundaries,
                 qboundaryMarkers=quantileBoundaries)
```

###  Quality checking model fit

```{r, visualization, eval=FALSE}


# check convergence
sapply(constrGMMfit$fit, function (x) x$convergence)
# visualize constrained fits with quantile drawn on histogram
trash = sapply(nzNormedMarkers, function(x) plot.gluster(constrGMMfit, marker = x, boundary = quantile(constrGMMfit$expressionX[,x], probs=quantileBoundaries[[x]][2,1], title=x) ) )

# comparing constrained and unconstrained fits
trash = sapply(nzNormedMarkers, function(x) hist_sr_constrast(constrGMMfit, unconstrGMMfit, marker=x, title=x) )


```

## Running with sub-batches

Sub-batches can be used if there are multiple regions per slide to assess the fit within each region. The model is fit at the slide level and then model fit is assessed by batch.

```{r subBatch, eval=FALSE}
# fitting on the same slide and getting model fit results by batch
batchConstrGMMfit = gluster(cells[[1]][,nzNormedMarkers], boundaryMarkers = boundaries,
                 qboundaryMarkers=quantileBoundaries, subBatch = cells[[1]]$sample_id)
# visualize by subBatch for the first markers
trash = sapply(unique(cells[[1]]$sample_id), function(x) plot.gluster(batchConstrGMMfit, subBatch = x, title=x) )

```

## Running gluster on multiple slides

We can use mclapply from the parallel package to run gluster on multiple slides.

```{r}
quantileBoundaries = list(matrix(c(0,.8, .9, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          NULL,
                          NULL,
                          NULL)
boundaries = list(matrix(c(0,.3, .5, 1), byrow = TRUE, nrow=2),
                  NULL,
                  NULL,
                  NULL,
                  matrix(c(0,.3, .45, Inf), byrow = TRUE, nrow=2),
                  matrix(c(0,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(0,.4, .45, Inf), byrow = TRUE, nrow=2),
                  NULL)
```

```{r, batchFit}
constrCfGMMbunch = groupGluster(cell[,nzNormedMarkers], slide=cell$slide_id, boundaryMarkers=boundaries, qboundaryMarkers = quantileBoundaries, n.cores = 40)

#trash = sapply(names(constrCfGMMbunch), function(x) plot.gluster(constrCfGMMbunch[[x]], marker = 1, boundary = quantile(constrGMMfit$expressionX[,1], probs=quantileBoundaries[[1]][2,1]), title = x ) )

library(reactable)
convergenceTab = sapply(constrCfGMMbunch, function(x) sapply(x$fit, function(y){ y$convergence}))
# the model converged for most slides
rowMeans(convergenceTab)
reactable(convergenceTab)


```

### Adjusting the boundary constraints

We are unlikely to get a biologically consistent model fit without applying constraints or evaluating the boundaries. To improve fit across all slides, we visually assess the fit and adjust our initial starting boundaries. This can be done by visualizing each histogram (time intensive) or visualizing the estimated parameters across all slides and assessing the fit.

```{r}
markerind = 3
invisible(capture.output(sapply(names(constrCfGMMbunch), function(x) plot(constrCfGMMbunch[[x]], marker = markerind, boundary = quantile(constrCfGMMbunch[[markerind]]$expressionX[,1], probs=quantileBoundaries[[markerind]][2,1]), title = x ) )))
```

After visualizing the fits we can adjust the boundaries.

```{r adjustedBoundaries}
quantileBoundaries2 = list(matrix(c(0,.8, .9, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.6, .8, 1), byrow = TRUE, nrow=2),
                          matrix(c(0,.7, .8, 1), byrow = TRUE, nrow=2),
                          NULL,
                          NULL,
                          NULL)
boundaries2 = list(matrix(c(0,.3, .5, 1), byrow = TRUE, nrow=2),
                  matrix(c(0,.3, .5, 1), byrow = TRUE, nrow=2),
                  matrix(c(0,.3, .5, 1), byrow = TRUE, nrow=2),
                  NULL,
                  matrix(c(0,.3, .45, Inf), byrow = TRUE, nrow=2),
                  matrix(c(0,.3, .5, Inf), byrow = TRUE, nrow=2),
                  matrix(c(0,.4, .45, Inf), byrow = TRUE, nrow=2),
                  NULL)
```
